---
title: R Workflow
author:
  - name: Frank Harrell
    url: https://hbiostat.org
date: last-modified
format:
  html:
    self-contained: true
    anchor-sections: true
    code-tools: true
    code-fold: false
    code-link: false
    fig-width: 6
    fig-height: 4.5
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    mainfont: Source Sans Pro
    theme: journal
    toc: true
    toc-depth: 3
    toc-location: left
    captions: true
    cap-location: margin
    table-captions: true
    tbl-cap-location: margin
    reference-location: margin
    include-in-header: /home/harrelfe/doc/html/ganalytics.html

comments:
  hypothesis: true

execute:
  warning: false
  message: false
---

<!-- To compile: quarto render rflow.qmd  -->

```{r include=FALSE}
source('~/r/rmarkdown/qbookfun.r')
```

[`r hypcomment`]{.aside}
This article outlines analysis project workflow that I've found to be efficient in making reproducible research reports using R with `Rmarkdown` and now `Quarto`.  I start by covering importing data, creating annotated analysis files, and running descriptive statistics on them with goals of understanding the data and their quality and completeness.  Functions in the `Hmisc` package are used to annotate data frames and data tables with labels and units of measurement, show metadata/data dictionaries, and to produce tabular and graphical statistical summaries.  Efficient and clear methods of recoding variables are given.  Several examples of processing and manipulating data using the `data.table` package are given, including some non-trivial longitudinal data computations.  General principles of data analysis are briefly surveyed and some flexible bivariate and 3-variable analysis methods are presented.  Finally, examples of caching results, parallel processing, and simulation are presented.  In the process several useful report writing methods are exemplified, including program-controlled creation of multiple report tabs.

::: {.callout-note collapse="true"}
# R Code Repositories Used in This Report
This report makes heavy use of the following R packages and Github repository:

* [`Hmisc` package](https://hbiostat.org/R/Hmisc) which contains functions for importing data, data annotation, summary statistics, statistical graphics, advanced table making, etc.
* [`data.table` package](https://www.rdocumentation.org/packages/data.table/versions/1.14.2) for data storage, retrieval, manipulation, munging, aggregation, merging, and reshaping
* [`ggplot2` package](https://ggplot2.tidyverse.org) for static graphics
* [`plotly` package](https://plotly.com/r) for interactive graphics
* [`rms` package](https://hbiostat.org/R/rms) for statistical modeling, validation, and presentation
* [`knitr` package](https://cran.r-project.org/web/packages/knitr) for running reproducible reports, and also providing `kable` and `kables` functions for simple html table printing
* [`rscripts` Github repository](https://github.com/harrelfe/rscripts) with utility functions such as
   + `dataChk` for data checking
   + `hashCheck` for checking if parent objects have changed so a slow analysis has to be re-run (i.e., talking control of caching)
   + `htmlList` to easily print vectors in a named list using `kable`
   + `htmlView`, `htmlViewx` for viewing data dictionaries/metadata in browser windows
   + `kabl` to make it easy to use `kable` and `kables` for making html tables
   + `maketabs` to automatically make multiple tabs in `Quarto` reports, each tab holding the output of an R object
   + `movStats` for statistics computed efficiently using `data.table` in moving overlapping windows of a continuous variable
   + `rsHelp` for viewing helps files for functions in `rscripts`
   + `runifChanged` which uses `hashCheck` to automatically re-run an analysis if needed, otherwise to retrieve previous results efficiently
   
Useful report building tools are gathered into the `rscripts` `reptools.r` script: `htmlList`, `kabl`, `maketabs`, `htmlView*`, `dataChk`, `scplot`.  These are all loaded when `reptools.r` is loaded.
The `rscripts` `Github` functions are accessed by the `Hmisc` function `getRs`, e.g.

```{r eval=FALSE}
require(Hmisc)
#ggplotlyr <- function(x, ...) x   ## ?????
getRs('reptools.r', put='source')
getRs('movStats.r', put='source')
```
   
All the available help files for functions in `rscripts` are at [hbiostat.org/R/rscripts](https://hbiostat.org/R/rscripts).  To view a help file for one of the functions in the `RStudio` `Viewer` pane use for example `rsHelp(movStats)` or `rsHelp(reptools)`.
:::

::: {.callout-note collapse="true"}
# R Basics
## Assignment Operator

You assign an R object to a value using the assignment operator `<-` or the equal sign.  `<-` is read as "gets".

```{r eval=FALSE}
x <- y
d <- read.csv('mydata.csv')
x = y
```

## Object Types

Everything in R is an object.  Some primitive types of objects in R are below.

| Type | Meaning |
|------|---------|
| integer | whole numbers |
| logical | values of `TRUE` or `FALSE` |
| double  | floating point non-whole numbers |
| character | character strings |
| function| code defining a function |

In the table below, objects of different shapes are described.
`rows` and `cols` refers to vectors of integers or logicals, or if the elements of the object are named, character strings.

| Type | Example |Values Retrieved By|
|------|---------|--------------------|
| scalar |  `x <- 3` | `x` |
| vector | `y <- c(1, 2, 5)` | `y[2]` (2), `y[2:3]` (2, 5), `y[-1]` (2, 5), `y[c(TRUE,FALSE,TRUE)]` (1) |
| named vector | `y <- c(a=1, b=2, d=5)` | `y[2]` (2), `y['b']` (2), `y[c('a','b')]` (1, 2) |
| matrix | `y <- cbind(1:3, 4:5)` | `y[rows,cols]`, `y[rows,]` (all cols), `y[,cols]` (all rows) |
| list | `x <- list(a='cat', b=c(1,3,7))` | `x$a` ('cat'), `x[[1]]` ('cat'), `x[['a']]` ('cat') |

Named vectors provide an extremely quick table lookup and recoding capability.

`list` objects are arbitrary trees and can have elements nested to any level.  You can have lists of lists or lists of data frames/tables.

Vectors can be of many different types when a `class` is added to them.  Two of the most common are `Date`s and `factor`s.  Character strings are handled very efficiently in R so there is not always a need to store categorical variables as `factor`s.  But there is one reason: to order levels, i.e., distinct variable values, so that tabular and graphical output will list values in a more logical order than alphabetic.  A factor variable has a `levels` _attribute_ added to it to accomplish this.  An example is `x <- factor(x, 1:3, c('cat', 'dog', 'fox'))` where the second argument `1:3` is the vector of possible numeric values `x` currently takes on (in order) and the three character strings are the corresponding `levels`.  Internally `factors` are coded as integers, but they print as character strings.

Rectangular data objects, i.e., when the number of rows is the same for every column (variable), can be represented by matrices, `data.frame`s, and `data.table`s.  In a matrix, every value is of the same type.  A `data.frame` or a `data.table` is an R `list` that can have mixtures of numeric, character, factor, dates, and other object types.  A `data.table` is also a `data.frame` but the converse isn't true.  `data.table`s are handled by the R `data.table` package and don't have row names but can be indexed, are much faster to process, and have a host of methods implemented for aggregation and other operations.  `data.frame`s are handled by base R.

## Subscripting

Examples of subscripting are given above. 
Subscripting via placement of `[]` after an object name is used for subsetting, and occasionally for using some elements more than once:

```{r subsc}
x <- c('cat', 'dog', 'fox')
x[2:3]
x[c(1, 1, 3, 3, 2)]
```

Subscripting a variable or a data frame/table by a vector of `TRUE/FALSE` values is a very powerful feature of R.  This is used to obtain elements satisfying one or more conditions:

```{r subcond}
x <- c(1, 2, 3, 2, 1, 4, 7)
y <- c(1, 8, 2, 3, 8, 9, 2)
x[y > 7]
```

The last line of code can be read as "values of `x` such that `y > 7`".

## Branching and If/Then

### Decisions Base on One Scalar Value

Common approaches to this problem are `if` and `switch`.

```{r ifs,eval=FALSE}
type <- 'semiparametric'
f <- switch(parametric     = ols(y ~ x),
            semiparametric = orm(y ~ x),
            nonparametric  = rcorr(x, y, type='spearman'),
            { z <- y / x
              c(median=median(z), gmean=exp(mean(log(z)))) } )
# The last 2 lines are executed for any type other than the 3 listed
f <- if(type == 'parametric')    ols(y ~ x)
  else
    if(type == 'semiparametric') orm(y ~ x)
  else
    if(type == 'nonparametric')  rcorr(x, y, type='spearman')
  else {
    z <- y / z
    c(median=median(z), gmean=exp(mean(log(z)))
  }
```

What is inside `if( )` must be a single scalar element that is evaluated to whether it's `TRUE` or `FALSE`.

### Series of Separate Decisions Over a Vector of Values

The `ifelse` or `data.table::fifelse` functions are most often used for this, but `data.table::fcase` is a little better.  Here's an example.

```{r ifelse}
x <- c('cat', 'dog', 'giraffe', 'elephant')
type <- ifelse(x %in% c('cat', 'dog'), 'domestic', 'wild')
type
require(data.table)
fcase(x %in% c('cat', 'dog'), 'domestic', default='wild')
```

## Functions

Even new R users can benefit from writing functions to reduce repetitive coding.  A function has _arguments_ and these can have default values for when the argument is not specified by the user when the function is called.  Here are some examples.  One line functions do not need to have their bodies enclosed in `{}`.

```{r runex}
cuberoot <- function(x) x ^ (1/3)
cuberoot(8)
g <- function(x, power=2) {
  u <- abs(x - 0.5)
  u / (1. + u ^ power)
}
g(3, power=2)
g(3)
# Function to make mean() drop missing values without our telling it
mn <- function(x) mean(x, na.rm=TRUE)
# Function to be used throughout the report to round fractional values 
# by a default amount (here round to 0.001)
rnd <- function(x) round(x, 3)
# edit the 3 the change rounding anywhere in the report
# The following simple function saves coding when you need to recode multiple
# variables from 0/1 to no/yes.
yn <- function(x) factor(x, 0:1, c('no', 'yes'))
```

:::

# Report Formatting

A state-of-the-art way to make reproducible reports is to use a statistical computing language such as R and its [`knitr`](https://cran.r-project.org/web/packages/knitr) package in conjunction with either [`RMarkdown`](https://rmarkdown.rstudio.com) or [`Quarto`](https://quarto.org), with the latter likely to replace the former.  Both of the report-making systems allow one to produce reports in a variety of formats including html, pdf, and Word.  Html is recommended because pages can be automatically resized to allow optimum viewing on devices of most sizes, and because html allows for interactive graphics and other interactive components.  Pdf is produced by converting `RMarkdown` or `Quarto`-produced markdown elements to $\LaTeX$.[Report formatting is very much enhanced by using variable attributes such as labels and units of measurement that are not considered in base R.  Methods for better annotating output using labels and units are given below.]{.aside}

This document can serve as a template for using R with `Quarto`; one can see the raw script by clicking on `Code` at the top right of the report.  When one has only one output format target, things are fairly straightforward except for the following.  To make use of specialized functions that produce html or $\LaTeX$ markup, one often has to put `results='asis'` in the code chunk header to keep the system from disturbing the generated html or $\LaTeX$ markup so that it will be typeset correctly in the final document.  This process works smoothly but creates one complication: if you print an object that produces plain text in the same code chunk, the system will try to typeset it in html or $\LaTeX$.  To prevent this from happening you either need to split the chunks into multiple chunks (some with `results='asis'` and some not) or you need to make it clear that parts of the output are to be typeset verbatim.  To do that a simple function `pr` can sense if `results='asis'` is in effect for the current chunk.  If so, the object is surrounded by the `markdown` verbatim indicator---three consecutive back ticks.  If not the object is left alone.  `pr` is defined in the `marksupSpecs$markdown$pr` object, so you can bring it to your session by copying into a local function `pr` as shown below, which has a chunk option `results='asis'` to show that verbatim output appears anyway.[If the argument `obj` to `pr` is a data frame or data table, variables will be rounded to the value given in the argument `dec` (default `dec=3`) before printing.  If you specify `inline=x` the object `x` is printed with `cat()` instead of `print()`.  `inline` is more for printing character strings.]{.aside}

::: {.column-margin}

An example of something that will not render correctly due to `results='asis'` being in the chunk header (needed for `html(...)`):

```{r eval=FALSE}
html(describe(d))
table(sex, country)
# use pr(obj=table(sex, country))
# to fix
```

:::

```{r,results='asis'}
require(Hmisc)
pr <- markupSpecs$markdown$pr
x <- (1:5)/7
pr('x:', x)
pr(obj=x)
pr(inline=paste(round(x,3), collapse=', '))
```

Instead of working to keep certain outputs verbatim you can use `knitr::kable()` to convert verbatim output to markdown.

## Multi-Output Format Reports

To allow one report to be used to render multiple output formats, especially html and pdf, it is helpful to be able to sense which output format is currently in play, and to use different functions or options to render output explicitly for the current format.  Here is how to create variables that can be referenced simply in code throughout the report, and to invoke the `plotly` graphics package if output is in html to allow interactivity.  A small function `ggp` is defined so that if you run any `ggplot2` output through it, the result will be automatically converted to `plotly` using the `ggplotly` function, otherwise it is left at standard static `ggplot2` output if html is not the output target.[See [this](https://hbiostat.org/ordinal) for examples of articles rendered in both html and PDF]{.aside}

```{r senseformat}
outfmt <- if(knitr::is_html_output ()) 'html'  else 'pdf'
markup <- if(knitr::is_latex_output()) 'latex' else 'html'
ishtml <- outfmt == 'html'
if(ishtml) require(plotly)
ggp <- if(ishtml) ggplotlyr else function(ggobject, ...) ggobject
# See below for more about ggplotlyr (a front end for ggplotly that can
# correct a formatting issue with hover text)
```

::: {.callout-note collapse="true"}
## Special Considerations For $\LaTeX$/PDF

The `Hmisc`, `rms`, and `rmsb` packages have a good deal of support for creating $\LaTeX$ output in addition to html.  They require some special $\LaTeX$ packages to be accessed.  In addition, if using any of `Quarto`'s nice features for making marginal notes, there is another $\LaTeX$ package to attach.  Below you'll find what needs to be added to the `yaml` prologue at the top of your script if using `Quarto`.  You have to modify `pdf-engine` to suit your needs.  I use `luatex` because it handles special unicode characters.  In the future (approximately July 2022) a bug in `Pandoc` will be fixed and you can put `links-as-notes: true` in the `yaml` header instead of redefining `href` and linking in `hyperref`.

```
format:
  html:
    self-contained: true
    . . .
  pdf:
    pdf-engine: lualatex
    toc: false
    number-sections: true
    number-depth: 2
    top-level-division: section
    reference-location: document
    listings: false
    header-includes:
      \usepackage{marginnote, here, relsize, needspace, setspace, hyperref}
      \renewcommand{\href}[2]{#2\footnote{\url{#1}}}
```

The `href` redefinition above turns URLs into footnotes if running $\LaTeX$.

There is one output element provided by `Quarto` that will not render correctly to $\LaTeX$: a marginal note using the markup `.column-margin`.  To automatically use an alternate in-body format, define a function that can be used for both typesetting formats.

```{r eval=FALSE}
mNote <- if(ishtml) '.column-margin'
  else
                    '.callout-note appearance="minimal"'
```

Then use <tt>r mNote</tt> enclosed in back ticks in place of the `.column-margin` callout for generality.

:::

::: {.callout-note collapse="true"}
## Using `Hmisc` Formatting for Variable Labels in Tables

As done with various `Hmisc` and `rms` package functions, one can capitalize on `Hmisc`'s special formatting of variable labels and units when constructing tables in $\LaTeX$ or html.  The basic constructs are shown in the code below.

```{r labtab,eval=FALSE}
# Retrieve a set of markup functions depending on typesetting format
# See below for definition of ishtml
specs    <- markupSpecs[[if(ishtml) 'html' else 'latex']]
# Hmisc markupSpecs functions create plain text, html, latex,
# markdown, or plotmath code
varlabel <- specs$varlabel  # retrieve an individual function
# Format text describing variable named x
# hfill=TRUE typesets units to be right-justified in label
# Use the following character string as a row label
# Default specifies the string to use if there is not label
# (usually taken as the variable name)
varlabel(label(x, default='x'), units(x), hfill=TRUE)
```

:::

# File Directory Structure

I have a directory for each major project, and put everything in that
directory (including data files) except for graphics files for
figures, which are placed in their own subdirectory underneath the
project folder.[With `Quarto` I specify that `html` files are to be self-contained, so there are no separate graphics files.]{.aside}  The directory name for the project includes key
identifying information, and files within that directory do not
contain project information in their names, nor do they contain dates,
unless I want to freeze an old version of an analysis script.

For multi-analyst projects or ones in which you want to capture the entire code history, having the project on `github` is worthwhile.

# Analysis File Creation

I typically create a compact analysis file in a separate R script
called `create.r` and have it produce compressed R binary `data.frame`
or `data.table` `.rds` files using `saveRDS(name, 'name.rds', compress='xz')`.  Then I have an analysis script named for example `a.qmd` (for [Quarto](https://quarto.org) reports) or `a.Rmd` (for RMarkdown reports) that starts with `d <- readRDS('name.rds')`.  [Templates for analysis reports are [here](http://hbiostat.org/rr/index.html#template), and a comprehensive report example may be found [here](https://hbiostat.org/R/Hmisc/markov).]{.aside}

When variables need to be recoded, have labels added or changed, or
have units of 
measurement added[Variable labels and units of measurement are used in special ways in my R packages.  This will show up in the `describe` and `contents` function outputs below and in axis labels for graphics.]{.aside}, I specify those using the `Hmisc` package [`upData`
function](https://www.rdocumentation.org/packages/Hmisc/versions/4.7-0/topics/upData).

To facilitate some operations requiring variable names to be quoted, define a function `.q` to quote them automatically. `.q` is like the `Hmisc` function `Cs` but also allows elements to be named.  It will be in `Hmisc` 4.7-1.

```{r quote}
.q <- function(...) {
  s <- sys.call()[-1]
  w <- as.character(s)
  n <- names(s)
  if(length(n)) names(w) <- n
  w
}

.q(a, b, c, 'this and that')
.q(dog=a, giraffe=b, cat=c)
```

Here is an `upData` example:

```{r eval=FALSE}
# Function to recode from atypical coding for yes/no in raw data
yn <- function(x) factor(x, 0:1, c('yes', 'no'))
d <-
  upData(d,
         rename = .q(gender=sex, any.event=anyEvent),
         posSE    = yn(posSE),
         newMI    = yn(newMI),
         newPTCA  = yn(newPTCA),
         newCABG  = yn(newCABG),
         death    = yn(death),
         hxofHT   = yn(hxofHT),
         hxofDM   = yn(hxofDM),
         hxofCig  = factor(hxofCig, c(0, 0.5, 1),
                           c('heavy', 'moderate', 'non-smoker')), 
         hxofMI   = yn(hxofMI),
         hxofPTCA = yn(hxofPTCA),
         hxofCABG = yn(hxofCABG),
         chestpain= yn(chestpain),
         anyEvent = yn(anyEvent),
         drop=.q(event.no, phat, mics, deltaEF,
                 newpkmphr, gdpkmphr, gdmaxmphr, gddpeakdp, gdmaxdp,
                 hardness),
         labels=c(
           bhr       = 'Basal heart rate',
           basebp    = 'Basal blood pressure',
           basedp    = 'Basal Double Product bhr*basebp',
           age       = 'Age',
           pkhr      = 'Peak heart rate',
           sbp       = 'Systolic blood pressure',
           dp        = 'Double product pkhr*sbp',
           dose      = 'Dose of dobutamine given',
           maxhr     = 'Maximum heart rate',
           pctMphr   = 'Percent maximum predicted heart rate achieved',
           mbp       = 'Maximum blood pressure',
           dpmaxdo   = 'Double product on max dobutamine dose',
           dobdose   = 'Dobutamine dose at max double product',
           baseEF    = 'Baseline cardiac ejection fraction',
           dobEF     = 'Ejection fraction on dobutamine', 
           chestpain = 'Chest pain', 
           ecg       = 'Baseline electrocardiogram diagnosis',
           restwma   = 'Resting wall motion abnormality on echocardiogram', 
           posSE     = 'Positive stress echocardiogram',
           newMI     = 'New myocardial infarction',
           newPTCA   = 'Recent angioplasty',
           newCABG   = 'Recent bypass surgery', 
           hxofHT    = 'History of hypertension', 
           hxofDM    = 'History of diabetes',
           hxofMI    = 'History of myocardial infarction',
           hxofCig   = 'History of smoking',
           hxofPTCA  = 'History of angioplasty',
           hxofCABG  = 'History of coronary artery bypass surgery',
           anyEvent  = 'Death, newMI, newPTCA, or newCABG'),
         units=.q(age=years, bhr=bpm, basebp=mmHg, basedp='bpm*mmHg',
           pkhr=mmHg, sbp=mmHg, dp='bpm*mmHg', maxhr=bpm,
           mbp=mmHg, dpmaxdo='bpm*mmHg', baseEF='%', dobEF='%',
           pctMphr='%', dose=mg, dobdose=mg)
         )

saveRDS(d, 'stressEcho.rds', compress='xz')

# Note that we could have automatically recoded all 0:1 variables
# if they were all to be treated identically:

for(x in names(d)) 
  if(all(d[[x]] %in% c(0,1))) d[[x]] <- yn(d[[x]])
```

::: {.panel-tabset}

## 

## External Metadata

Sometimes metadata comes from a separate source.  Suppose you imported a data frame `d` but have also imported a data frame `m` containing metadata: the same variable names in `d` (variable `name`) plus fields `label`, `units`, and `comment`.  Dataset `m` can contain variables not currently being used.  To add the labels and units into `d` and to store comments separately, use the following example.

```{r eval=FALSE}
n <- names(d)
i <- n %nin% m$name
if(any(i)) cat('The following variables have no metadata:',
               paste(n[i], collapse=', '), '\n')
vcomment        <- m$comment
names(vcomment) <- m$name
mn              <- subset(m, name %in% n)
labs            <- mn$label
un              <- mn$units
names(labs)     <- names(un) <- mn$name
d <- upData(d, labels=labs, units=un)
```

To look up the `comment` for a variable at any time use e.g. `vcomment['height']`.  All comments were saved in vector `vcomment` in case the metadata dictionary `m` defined variables that are not in `d` but were to be imported later in the script.  Comments for multiple variables can be looked up using e.g. `vcomment[.q(height, weight, age)]`.

::: {.column-margin}
If you want to look up a variable's comment without having to quote its name use the following:

```{r vcom,eval=FALSE}
vcom <- function(...)
  vcomment[as.character(sys.call()[-1])]
# Example usage: vcom(age,sbp,dbp)
```

:::

## CSV

The built-in function in R for reading `.csv` files is [`read.csv`](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table).  The `Hmisc` package has a function [`csv.get`](https://www.rdocumentation.org/packages/Hmisc/versions/4.7-0/topics/csv.get) which calls `read.csv` but offers some enhancements in variable naming, date handling, and reading variable labels from a specified row number.  Illegal characters in variable names are changed to periods, and by default underscores are also changed to periods.  If any variable names are changed and the `labels` argument is not given, original variable names are stored in the variable `label` attributes.

The [`fread`](https://www.rdocumentation.org/packages/data.table/versions/1.14.2/topics/fread) function in the `data.table` package is blazing fast for reading large files and offers a number of options.  `csv.get` uses it if `data.table` is installed on the system.

## REDCap

If reading data exported from `REDCap` that are placed into the
project directory I run the following to get rid of duplicate
(`factor` and non-`factor` versions of variables `REDCap` produces)
variables and automatically convert dates to `Date` variables:

```
require(Hmisc)
getRs('importREDCap.r', put='source')  # source() code to define function
mydata <- importREDCap()  # by default operates on last downloaded export
saveRDS(mydata, 'mydata.rds', compress='xz')
```

When file names are not given to `importREDCap` the function looks for
the latest created `.csv` file and `.R` file with same prefix and uses those.
See [this](http://hbiostat.org/bbr/md/r.html) for more information.

## SAS, Stata, SPSS

SAS, Stata, and SPSS binary files are converted to R `data.frame`s using the R [haven package](https://haven.tidyverse.org).  Here's an example:

```{r haven, eval=FALSE}
require(haven)   # after you've installed the package
d <- read_sas('mydata.sas7bdat')
d <- read_xpt('mydata.xpt')        # SAS transport files
d <- read_dta('mydata.dta')        # Stata
d <- read_sav('mydata.sav')        # SPSS
d <- read_por('mydata.por')        # Older SPSS files
```

These import functions carry variable labels into the data frame and convert dates and times appropriately.  Character vectors are not converted to factors.

## Multiple Files

One of the most important principles to following in programming data analyses is to not do the same thing more than once.  Repetitive code wastes time and is harder to maintain.  One example of avoiding repetition is in reading a large number of files in R.  If the files are stored in one directory and have a consistent file naming scheme (or you want to import every `.csv` file in the directory), one can avoid naming the individual files.  The results may be stored in an R `list` that has named elements, and there are many processing tasks that can be automated by looping over this list.

In the following example assume that all the data files are `.csv` files in the current working directory, and they all have names of the form `xz*.csv`.  Let's read all of them and put each file into a data frame named by the characters in front of `.csv`.  These data frames are stuffed into a `list` named `X`.  The `Hmisc` `csv.get` function is used to read the files, automatically translating dates to `Date` variables, and because `lowernames=TRUE` is specified, variable names are translated to lower case.  There is an option to fetch variable labels from a certain row of each `.csv` file but we are not using that.

```{r eval=FALSE}
files <- list.files(pattern='xz.*.csv')  # vector of qualifying file names
# Get base part of *.csv
dnames <- sub('.csv', '', files)
X <- list()
i <- 0
for(f in files) {
  cat('Reading file', f, '\n')
  i <- i + 1
  d <- csv.get(f, lowernames=TRUE)
  # To read SAS, Stata, SPSS binary files use a haven function instead
  # To convert to data.table do setDT(d) here
  X[[dnames[i]]] <- d
}
saveRDS(X, 'X.rds', compress='xz')  # Efficiently store all datasets together
```

To process one of the datasets one can do things like `summary(X[[3]])` or `summary(X$baseline)` where the third dataset stored was named `baseline` because it was imported from `baseline.csv`.

Now there are many possibilities for processing all the datasets at once such as the following.

```{r processmany,eval=FALSE}
k   <- length(X)          # number of datasets
nam <- lapply(X, names)   # list with k elements, each is a vector of names
# Get the union of all variable names used in any dataset
sort(unique(unlist(nam))) # all the variables appearing in any dataset
# Get variable names contained in all datasets
common <- names(X[[1]])
for(i in 2 : k) {
  common <- intersect(common, names(X[[i]]))
  if(! length(common)) break  # intersection already empty
}
sort(common)
# Compute number of variables across datasets
nvar <- sapply(X, length)  # or ncol
# Print number of observations per dataset
sapply(X, nrow)
# For each variable name count the number of datasets containing it
w <- data.table(dsname=rep(names(X), nvar), vname=unlist(nam))
w[, .N, keyby=vname]
# For each variable create a comma-separated list of datasets
# containing it
w[, .(datasets=paste(sort(dsname), collapse=', ')), keyby=vname]
# For datasets having a subject ID variable named id compute
# the number of unique ids
uid <- function(d) if('id' %in% names(d)) length(unique(d$id)) else NA
sapply(X, uid)
# To repeatedly analyze one of the datasets, extract it to a single data frame
d <- X$baseline
describe(d)
```

## Post Import

The `Hmisc` package `cleanup.import` function improves imported data storage in a number of ways including converting double precision variables to `integer` when originally `double` but not containing fractional values (this halves the storage requirement).  `Hmisc::upData` is the go-to function for annotating data frames/tables, renaming variables, and dropping variables.  `Hmisc::dataframeReduced` removes problematic variables, e.g., those with a high fraction of missing values or that are binary with very small prevalence.

:::

## Variable Naming

I prefer short but descriptive variable names.  As exemplified above, I use variable `label`s and `units` to provide more information.  For example I wouldn't name the age variable `age.at.enrollment.yrs` but would name it `age` with a label of `Age at Enrollment` and with `units` of `years`.  Short, clear names unclutter code, especially formulas in statistical models.  One can always fetch a variable label while writing a program (e.g., typing `label(d$age)` at the console) to check that you have the right variable (or put the data dictionary in a window for easy reference, as shown below).  `Hmisc` package graphics and table making functions such as `summaryM` and `summary.formula` specially typeset `units` in a smaller font.


## Data Dictionary

The `Hmisc` package `contents` function will provide a concise data dictionary.    Here is an example using the permanent version[The permanent version of `stressEcho` coded binary variables as 0/1 instead of N/Y.]{.aside} of the dataset created above, which can be accessed with the `Hmisc` `getHdata` function.  The top of the `contents` output has the number of levels for `factor` variables hyperlinked.  Click on the number to go directly to the list of levels for that variable.

```{r getdat}
require(Hmisc)
getHdata(stressEcho)
d <- stressEcho
```

::: {.callout-note collapse="true"}
# Data Dictionary

```{r ddict2,results='asis'}
html(contents(d), levelType='table')
```
:::

You can write the text output of `contents` into a text file in your current working directory, and click on that file in the `RStudio` `Files` window to create a new tab in the editor panel where you can view the data dictionary at any time.  This is especially helpful if you need a reminder of variable definitions that are stored in the variable labels.  Here is an example where the formatted data dictionary is saved.  [Users having the `xless` system command installed can pop up a `contents` window at any time by typing `xless(contents(d))` in the console. `xless` is in `Hmisc`.]{.aside}


```{r,eval=FALSE}
capture.output(contents(d), file='contents.txt')
```

Or put the html version of the data dictionary into a small browser window to which you can refer at any point in analysis coding.

```{r eval=FALSE}
cat(html(contents(d)), file='contents.html')
browseURL('contents.html', browser='vivaldi -new-window')
```


`RStudio` provides a nice way to do this, facilitated by the `htmlView` function which may be fetched from `Github` using the `Hmisc` `getRs` function as shown below.  `htmlView` takes any number of objects for which an `html` method exists to render them.  They are rendered in the `RStudio` `Viewer` pane.  If you are running outside `RStudio`, your default browser will be launched instead.[Occasionally `RStudio Viewer` will drop its arrow button making it impossible to navigate back and forth to different `html` outputs.<br>Code for `htmlView` and `htmlViewx` may be viewed in [`reptools.r`](https://github.com/harrelfe/rscripts/blob/master/reptools.r).]{.aside}

```{r eval=FALSE}
getRs('reptools.r', put='source')
# reptools.r defines htmlView, htmlViewx, kabl, maketabs, dataChk
htmlView(contents(d))
```

In some cases it is best to have a browser window open to the full descriptive statistics for a data table/frame (see below; the `describe` function also shows labels, units, and levels).  [For either approach it would be easy to have multiple tabs open, one tab for each of a series of data tables, or use `htmlView`.]{.aside}

To be able to have multiple windows open to see information about datasets it is advantageous to open an external browser window.  The `htmlViewx` function defined when you read `htmlView.r` will by default open a Vivaldi browser window with the first output put in a new window and all subsequent objects displayed as tabs within that same window.  This behavior can be controlled with the `tab` argument, and you can use a different browser by issuing for example `options(vbrowser='firefox')`.[In Windows you may have to specify a full path and `firefox.exe`.  The tab names will not be correct until `Hmisc` 4.7-1 appears.]{.aside}  As an example suppose that two datasets were read from the `hbiostat.org/data` data repository, and the data dictionary and descriptive statistics for both datasets were to be converted to `html` and placed in an external browser window for the duration of the R session.

```{r vbrowse,eval=FALSE}
getHdata(support)
getHdata(support2)
htmlViewx(contents(support ), describe(support ),
          contents(support2), describe(support2))
```

A screenshot of the result is [here](https://hbiostat.org/img/htmlViewx.png).

# Data Checking

Besides useful descriptive statistics exemplified below, it is important to flag suspicious values in an automated way.  Since checking multiple columns may involve a large number of R expressions to run to classify observations as suspicious, let's automate the process somewhat by specifying a vector of expressions.  Then we have R "compute on the language" to parse the expressions for finding observations to flag, and for printing.  This is done by the [`dataChk`](https://github.com/harrelfe/rscripts/blob/master/reptools.r) function in `Github`.

The following code results in separate output for each individual data check, in separate `Quarto` tabs.  The dataset does not have a subject ID variable so let's create one, and also add a `site` variable to print.  Arguments are specified to `dataChk` so that no tab is produced for a condition that never occurred in the data, and a tab is produced showing all data flags, sorted by `id` and `site`.

```{r dcheck,results='asis'}
getRs('reptools.r', put='source')  # Define dataChk, maketabs
require(data.table)
w <- d
setDT(w)
w[, id := 1 : .N]
set.seed(1)
w[, site := sample(LETTERS[1:6], .N, replace=TRUE)]
checks <- expression(
  age < 30 | age > 90,
  gender == 'female' & maxhr > 170,
  baseEF %between% c(72, 77),
  baseEF > 77,
  baseEF > 99,
  sbp > 250 & maxhr < 160)
dataChk(w, checks, id=c('id', 'site'),
        omit0=TRUE, byid=TRUE, html=TRUE)
```

# Descriptive Statistics

The `Hmisc` `describe` function is my main tool for getting initial descriptive statistics and quality controlling the data in a univariate fashion.  Here is an example.  The `Info` index is a measure of the information content in a numeric variable relative to the information in a continuous numeric variable with no ties.  A very low value of `Info` will occur when a highly imbalanced variable is binary.  Clicking on `Glossary` on the right will pop up a browser window with a more in-depth glossary of terms used in `Hmisc` package output.  It links to `hbiostat.org/R/glossary.html` which you can link  from your reports that use `Hmisc`.[`Info` comes from the [approximate formula](http://hbiostat.org/bib/r2.html) for the variance of a log odds ratio for a proportional odds model/Wilcoxon test, due to Whitehead.<br><a href="https://hbiostat.org/comment.html" target="popup" onclick="window.open('https://hbiostat.org/R/glossary.html#describe', 'popup', 'width=450,height=600'); return false;"><small>Glossary</small></a>]{.aside}

`Gmd` in the output stands for Gini's mean difference---the mean absolute difference over all possible pairs of different observations.  It is a very interpretable measure of dispersion that is more robust than the standard deviation.

::: {.callout-note .column-page collapse="true"}
# `describe` Output
```{r desc,results='asis'}
w <- describe(d)
html(w)
```
:::

```{r eval=FALSE}
# To create a separate browser window:
cat(html(w), file='desc.html')
browseURL('desc.html', browser='firefox -new-window')
```

Better, whether using `RStudio` or not:

```{r eval=FALSE}
htmlView(w, contents(d))  # or htmlView(describe(d1), describe(d2), ...)
# Use htmlViewx to use an external browser window (see above)
```

There is also a `plot` method for `describe` output.  It produces two graphics objects: one for categorical variables and one for continuous variables.  The default is to use `ggplot2` to produce static graphics.

```{r pdesc}
p <- plot(w, bvspace=2.5)
```

::: {.panel-tabset}
## Categorical Variables
```{r pdesccat}
p$Categorical
```

## Continuous Variables
```{r pdesccon}
p$Continuous
```

:::

To make creation of tabs in `Quarto` easier, the `maketabs` function loops through the elements of a named list and outputs each element into a separate `Quarto` tab.  A `wide` argument is used to expand the width of the output outside the usual margins.  An `initblank` argument creates a first tab that is empty.  This allows one to show nothing until one of the other tabs is clicked.  [This function was defined by the earlier `getRs('reptools.r', put='source')` command and its source code may be viewed [here](https://github.com/harrelfe/rscripts/blob/master/reptools.r).]{.aside}

```{r maketabs,results='asis'}
maketabs(p)
```

By specifying `grType` option you can instead get `plotly` graphics that use hover text to show more information, especially when hovering over the leftmost dot or tick mark for a variable.

```{r pldesc,results='asis'}
options(grType='plotly')
p <- plot(w, bvspace=2.5)
maketabs(p, wide=TRUE)
```

See [this](http://hbiostat.org/R/Hmisc/examples.html) for other `Hmisc` functions for descriptive graphics and tables, especially for stratified descriptive statistics for categorical variables.  The `summaryM` function prints a tabular summary of a mix of continuous and categorical variables.  Here is an example where stratification is by history of myocardial infarction (MI).  [The `initblank` option to the `maketabs` function creates the initially opened tab, which has no content, thereby leaving it to the user to click one of the other tabs before output is shown.]{.aside}

```{r summaryM,results='asis'}
require(data.table)
setDT(d)   # turn d into a data table
# tables() with no arguments will give a concise summary of all active data tables
w <- d
w[, hxofMI := factor(hxofMI, 0 : 1, c('No history of MI', 'History of MI'))]
vars <- setdiff(names(d), 'hxofMI')
form <- as.formula(paste(paste(vars, collapse='+'), '~ hxofMI'))
print(form)
s <- summaryM(form, data=d, test=TRUE)
w <- list('Table 1'                   = html(s, exclude1=TRUE, npct='both', digits=3, middle.bold=TRUE),
          'Categorical Variable Plot' = plot(s, which='categorical', vars=1 : 4, height=600, width=1000),
          'Continuous Variable Plot'  = plot(s, which='continuous',  vars=1 : 4))
# initblank=TRUE: Put an empty tab first so that nothing initially displays
maketabs(w, wide=TRUE, initblank=TRUE)
```

Semi-interactive stratified spike histograms are also useful descriptive plots.  These plots also contain a superset of the quantiles used in box plots, and the legend is clickable, allowing any of the statistical summaries to be turned off.

```{r spikeh}
d[, histboxp(x=maxhr, group=ecg, bins=200)]
```

# Data Manipulation and Aggregation

The [`data.table`](https://hbiostat.org/R/data.table) package provides a concise, consistent syntax for managing simple and complex data manipulation tasks, and it is extremely efficient for large datasets.  One of the best organized tutorials is [this](https://riptutorial.com/data-table), and a cheatsheet for data transformation is [here](https://raw.githubusercontent.com/rstudio/cheatsheets/master/datatable.pdf).  A master cheatsheet for `data.table` is [here](https://rdrr.io/cran/data.table/man/data.table.html) from which the general syntax below is taken, where `DT` represents a data table.

```
   DT[ i,  j,  by ] # + extra arguments
        |   |   |
        |   |    -------> grouped by what?
        |    -------> what to do?
         ---> on which rows?
```

Data tables are also data frames so work on any method handling data frames.  But data tables do not contain `rownames`.

Several `data.table` examples follow.  I like to hold the current dataset in `d` to save typing.  Some basic operations on data tables are:

```{r eval=FALSE}
d[2]                  # print 2nd row
d[2:4]                # print rows 2,3,4
d[y > 2 & z > 3]      # rows satisfying conditions
d[, age]              # retrieve one variable
d[, .(age, gender)]   # make a new table with two variables
i <- 2; d[, ..i]      # get column 2
v <- 'age'; d[, ..v]  # get variable named by contents of v
d[.N]                 # last row
d[, .N, keyby=age]    # number of rows for each age, sorted
d[1:10, bhr:pkhr]     # first 10 rows, variables bhr - pkhr
d[1:10, !(bhr:pkhr)]  # all but those variables
d[, 2:4]              # get columns 2-4
```

## Analyzing Selected Variables and Subsets
```{r dtsimple,results='asis'}
d[, html(describe(age))]
d[, html(describe(~ age + gender))]
d[gender == 'female', html(describe(age))]   # analyze age for females
html(describe(d[, .(age, gender)], 'Age and Gender Stats'))
```

```{r dtsimple2}
# Separate analysis by female, male.  Use keyby instead of by to sort the usual way.
d[, print(describe(age, descript=gender)), by=gender]

# Compute mean and median age by gender
d[, .(Mean=mean(age), Median=median(age)), by=gender]
# To create a new subset
w <- d[gender == 'female' & age < 70, ]
```

## Adding or Changing Variables

With `data.table` you can create a new data table with added variables, or you can add or redefine variables in a data table in place.  The latter has major speed and memory efficiency implications when processing massive data tables.  Here `d` refers to a different data table from the one used above.

```{r eval=FALSE}
# Rename a variable
setnames(d, c('gender', 'height'),
            c(  'sex',      'ht'))
# Easier:
setnames(d, .q(gender, height),
            .q(   sex,     ht))

# Or
ren <- .q(gender=sex, height=ht)
setnames(d, names(ren), ren)

# Or
rename <- function(x, n) setnames(x, names(n), n)
rename(d, .q(gender=sex, height=ht))

# For all variables having baseline_ at the start of their names, remove it
n <- names(d)
setnames(d, n, sub('^baseline_', '', n))   # ^ = at start; use $ for at end

# For all variables having baseline_ at the start of their names, remove it
# and add Baseline to start of variable label
n    <- names(d)
n    <- n[grepl('^baseline_', n)]
ren  <- sub('^baseline_', '', n); names(ren) <- n
# Fetch vector of labels for variables whose names start with baseline_
labs <- sapply(d, label)[n]   # label() result is "" if no label
labs <- paste('Baseline', labs)
d    <- updata(d, rename=ren, labels=labs)

# Change all names to lower case
n <- names(d)
setnames(d, n, tolower(n))

# Abbreviate names of all variables longer than 10 characters
# abbreviate() will ensure that all names are unique
n <- names(d)
setnames(d, n, abbreviate(n, minlength=10))

# For any variables having [...] or (...) in their labels, assume these
# are units of measurement and move them from the label to the units
# attribute
d <- upData(d, moveUnits=TRUE)

# Add two new variables, storing
result in a new data table
z <- d[, .(bmi=wt / ht ^ 2, bsa=0.016667 * sqrt(wt * ht))]

# Add one new variable in place
d[, bmi := wt / ht ^ 2]

# Add two new variables in place
d[, `:=`(bmi = wt / ht ^ 2, bas=0.016667 * sqrt(wt * ht))]
d[, .q(bmi, bsa) := .(wt / ht ^ 2, 0.016667 * sqrt(wt * ht))]

# Compute something requiring a different formula for different types
# of observations
d[, htAdj := ifelse(sex == 'male', ht, ht * 1.07)]  # better" use fifelse in data.table
d[, htAdj := ht * ifelse(sex == 'male', 1, 1.07)]
d[, htAdj := (sex == 'male') * ht + (sex == 'female') * ht * 1.07]
d[sex == 'male',   htAdj := ht]
d[sex == 'female', htAdj := ht * 1.07]
d[, htAdj := fcase(sex == 'male',   ht,          # fcase is in dta.table
                   sex == 'female', ht * 1.07)]
d[, htAdj := fcase(sex = 'female', ht * 1.07, default = ht)]

# Add label & optional units (better to use upData which works on data tables)
adlab <- function(x, lab, un='') {
  label(x) <- lab
  if(un != '') units(x) <- un
  x
}
d[, maxhr := adlab(maxhr, 'Maximum Heart Rate', '/m')]

# Delete a variable (or use upData)
d[, bsa := NULL]

# Delete two variables
d[, `:=`(bsa=NULL, bmi=NULL)]
d[, .q(bsa, bmi) := NULL]
```

## Recoding Variables

```{r, eval=FALSE}
# Group levels a1, a2 as a & b1,b2,b3 as b
d[, x := factor(x, .q(a1,a2,b1,b2,b3),
                   .q( a, a, b, b, b))]
# Regroup an existing factor variable
levels(d$x) <- list(a=.q(a1,a2), b=.q(b1,b2,b3))
# or
d <- upData(d, levels=list(x=list(a=.q(a1,a2), b=.q(b1,b2,b3))))
# Or manipulate character strings
d[, x := substring(x, 1, 1)]   # replace x with first character of levels
# or
levels(d$x) <- substring(levels(d$x), 1, 1)

# Recode a numeric variable with values 0, 1, 2, 3, 4 to 0, 1, 1, 1, 2
d[, x := 1 * (x %in% 1:3) + 2 * (x == 4)]
d[, x := fcase(x %in% 1:3, 1,          # fcase is in data.table
               x == 4,     2)]
d[, x := fcase(x %between% c(1,3), 1,
               x    ==       4,    2)]  # %between% is in data.table

# Recode a series of conditions to a factor variable whose value is taken
# from the last condition that is TRUE using Hmisc::score.binary
# Result is a factor variable unless you add retfactor=FALSE
d[, x := score.binary(x1 == 'symptomatic',
                      x2 %in% .q(stroke, MI),
                      death)]
# Same but code with numeric points
d[, x := score.binary(x1 == 'symptomatic',
                      x2 %in% .q(stroke, MI),
                      death,  # TRUE/FALSE or 1/0 variable
                      points=c(1,2,10))]
# Or just reverse the conditions and use fcase which stops at the first
# condition met
d[, x := fcase(death,                  'death',        # takes precedence
               x2 %in% .q(stroke, MI), 'stroke/MI',    # takes next precedence
               x1 == 'symptomatic',    'symptomatic',
               default =               'none')]

# Recode from one set of character strings to another using named vector
# A named vector can be subscripted with character strings as well as integers
states <- c(AL='Alabama', AK='Alaska', AZ='Arizona', ...)
# Could also do:
#  states <- .q(AL=Alabama, AK=Alaska, AZ=Arizona, ..., NM='New Mexico', ...) 
# or do a merge for table lookup (see later)
d[, State := states[state]]
# states are unique, state can have duplicates and all are recoded
d[, State := fcase(state == 'AL', 'Alabama',  state='AK', 'Alaska', ...)]

# Recode from integers 1, 2, ..., to character strings
labs <- .q(elephant, giraffe, dog, cat)
d[, x := labs[x]]

# Recode from character strings to integers
d[, x := match(x, labs)]
d[, x := fcase(x == 'elephant', 1, 
               x == 'giraffe',  2,
               x == 'dog',      3,
               x == 'cat',      4)]
```

As an example of more complex hierarchical recoding let's define codes in a nested `list`.

```{r hrecode}
a <- list(plant =
            list(vegetable = .q(spinach, lettuce, potato),
                 fruit     = .q(apple, orange, banana, pear)), 
          animal =
            list(domestic  = .q(dog, cat, horse),
                 wild      = .q(giraffe, elephant, lion, tiger)) )
a
a <- unlist(a)
a
# Pick names of unlist'ed elements apart to define kingdom and type
n <- sub('[0-9]*$', '', names(a))  # remove sequence numbers from ends of names
# Names are of the form kingdom.type; split at .
s       <- strsplit(n, '.', fixed=TRUE) 
kingdom <- sapply(s, function(x) x[1])
type    <- sapply(s, function(x) x[2])
# or:   (note \\. is escaped . meaning not to use as wild card)
#        .* = wild card: any number of characters
# kingdom <- sub('\\..*', '', n)  # in xxx.yyy remove .yyy
# type    <- sub('.*\\.', '', n)  # in xxx.yyy remove xxx.
names(kingdom) <- names(type) <- a
w <- data.table(kingdom, type, item=a, key=c('kingdom', 'item'))
w
# Example table lookups
cat(kingdom['dog'], ':', type['dog'], '\n')
kingdom[.q(dog, cat, spinach)]
type   [.q(dog, cat, giraffe, spinach)]
# But what if there is a plant named the same as an animal?
# Then look up on two keys
w[.('animal', 'lion'), type]
```

# Computing Summary Statistics

Many applications can use the automatically created `data.table` object `.SD` which stands for the data table for the current group being processed.  If `.SDcols` were not specified, all variables would be attempted to be analyzed.  Specify a vector of variable names as `.SDcols` to restrict the analysis.  If there were no `by` variable(s), `.SD` stands for the whole data table.

```{r ss}
# Compute the number of distinct values for all variables
nd <- function(x) length(unique(x))
d[, sapply(.SD, nd)]
# Same but only for variables whose names contain hx and either D or M
d[, sapply(.SD, nd), .SDcols=patterns('hx', 'D|M')]

# Compute means on all numeric variables
mn <- function(x) mean(x, na.rm=TRUE)
d[, lapply(.SD, mn), .SDcols=is.numeric]
# Compute means on all numeric non-binary variables
nnb <- function(x) is.numeric(x) && length(unique(x)) > 2
d[, lapply(.SD, mn), .SDcols=nnb]

# Print frequency tables of all categorical variables with > 2 levels
cmult <- function(x) ! is.numeric(x) && length(unique(x)) > 2
tab <- function(x) {
  z <- table(x, useNA='ifany')
  paste(paste0(names(z), ': ', z), collapse=', ')
}
d[, lapply(.SD, tab), .SDcols=cmult]
```

Tabulate all variables having between 3 and 10 distinct values and create a side effect when `data.table` is running that makes the summarization function `tab` store all values and frequencies in a growing list `Z` so that `kable` can render a `markdown` table after we pad columns to the maximum length of all columns (maximum number of distinct values). 

```{r diffreq}
# Using <<- makes data.table have a side effect of augmenting Z and
# Align in the global environment
tab <- function(x) {
  z <- table(x, useNA='ifany')
  i <- length(Z)
  Z[[i+1]] <<- names(z)
  Z[[i+2]] <<- as.vector(z)
  Align <<- c(Align, if(is.numeric(x)) 'r' else 'l', 'r')
  length(z)
}
discr <- function(x) { i <- length(unique(x)); i > 2 & i < 11 }
#                                       or i %between% c(2,11)
Z    <- list(); Align <- character(0)
w    <- d[, lapply(.SD, tab), .SDcols=discr]
maxl <- max(w)
# Pad shorter vectors with blanks
Z <- lapply(Z, function(x) c(x, rep('', maxl - length(x))))
Z <- do.call('cbind', Z)  # combine all into columns of a matrix
colnames(Z) <- rep(names(w), each=2)
colnames(Z)[seq(2, ncol(Z), by=2)] <- 'Freq'
knitr::kable(Z, align=Align)
```

A better approach is to let the `kables` function put together a series of separate `markdown` tables of different sizes.  By using the "updating `Z` in the global environment" side effect we are able to let `data.table` output any type of objects of non-conformable dimensions over variables (such as frequency tabulations).

```{r tabalt}
tab <- function(x) {
  z <- table(x, useNA='ifany')
  i <- length(Z)
  w <- matrix(cbind(names(z), z), ncol=2,
              dimnames=list(NULL, c(vnames[i+1], 'Freq')))
  Z[[i+1]] <<- knitr::kable(w, align=c(if(is.numeric(x)) 'r' else 'l', 'r'))
  length(z)
}
discr <- function(x) { i <- length(unique(x)); i > 2 & i < 11 }
Z      <- list()
vnames <- names(d[, .SD, .SDcols=discr])
w      <- d[, lapply(.SD, tab), .SDcols=discr]
knitr::kables(Z)
```

Use a similar side-effect approach to get separate `html` `describe` output by `gender`.

```{r bydesc,results='asis'}
g <- function(x, by) {
  Z[[length(Z) + 1]] <<- describe(x, descript=paste('age for', by))
  by
}
Z <- list()
by <- d[, g(age, gender), by=gender]
# Make Z look like describe() output for multiple variables
class(Z) <- 'describe'
attr(Z, 'dimensions') <- c(nrow(d), nrow(by))
attr(Z, 'descript') <- 'Age by Gender'
html(Z)
```

```{r morestats}
# Compute a 1-valued statistic on multiple variables, by cross-classification
# of two variables.  Do this on a subset.  .SDcols=a:b uses variables in order
# Use keyby instead of by to order output the usual way
d[age < 70, lapply(.SD, mean), keyby=.(gender, newMI), .SDcols=pkhr:dp]

# Compute multiple statistics on one variable
# Note: .N is a special variable: count of obs for current group
d[, .(Max=max(bhr), Min=min(bhr), Mean=mean(bhr), N=.N), by=.(gender, newMI)]
# Same result another way
g <- function(x) list(Max=max(x), Min=min(x), Mean=mean(x), N=length(x))
d[, g(bhr), by=.(gender, newMI)]  # if g returned a vector instead, use as.list(g(bhr))

d[, as.list(quantile(bhr)), by=gender]

# Compute mean bhr by quintiles of age using Hmisc::cut2
# Bad statistical practice; use scatterplot + smoother instead
d[, .(Mean=mean(bhr)), keyby=.(fifth=cut2(age, g=5))]

# Compute multiple statistics on multiple variables
d[, lapply(.SD, quantile), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# Similar but put percentile number in front of statistic value
# Do only quartiles
g <- function(x) {
  z <- quantile(x, (1:3)/4, na.rm=TRUE)
  paste(format(names(z)), format(round(z)))
}
d[, lapply(.SD, g), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# To have more control over labeling and to have one row per sex:
g <- function(x) {
  s <- sapply(x, quantile, na.rm=TRUE)  # compute quantiles for all variables -> matrix
  h <- as.list(s)  # vectorizes first
  # Cross row names (percentile names) with column (variable) names
  # paste(b, a) puts variable name in front of percentile
  names(h) <- outer(rownames(s), colnames(s), function(a, b) paste(b, a))
  h
}
d[, g(.SD), by=gender, .SDcols=.q(bhr, pkhr, sbp)]
# Restrict to variables bhr - basedp in order columns created in data table
d[, g(.SD), by=gender, .SDcols=bhr : basedp]
# Can put ! in front of a sequence of variables to do the opposite

# To add duplicated means to raw data use e.g.
# d[, Mean := mean(x), by=sex]
```

## Summary Statistics With Marginal Summaries

The `cube` function in the `data.table` package will compute all possible marginal statistics.  When there is only one `by` variable, the overall statistic is computed in addition to compute stratified values.  When a dimension is being marginalized over, the value of the `by` variable for that dimension will be `NA`.

```{r cube}
mn  <- function(x) as.double(mean(x, na.rm=TRUE))
# as.double ensures consistency of numeric type across groups
Nna <- function(x) sum(! is.na(x))
cube(d, .(Meanbhr = mn(bhr), N = Nna(bhr)), by='gender', id=TRUE)
cube(d, .(Meanbhr = mn(bhr), N = Nna(bhr)), by=.q(gender, hxofMI), id=TRUE)
# id=TRUE creates output variable grouping to detail level of marginalization
# It is a binary representation, e.g. if by has 3 variables and a row
# is marginalizing over the first and third variables,
# grouping=binary 101 = 5
# Use groupingsets() to control the marginalizations
# Example: marginalize only one variable at a time
groupingsets(d, .(Meanbhr = mn(bhr), N=Nna(bhr)), by=.q(gender, hxofMI),
             sets=list('gender', 'hxofMI'), id=TRUE)
```

# Merging Data

Consider a baseline dataset `b` and a longitudinal dataset `L`, with subject ID of `id`. [For more information see [this](https://hbiostat.org/R/data.table), [this](https://rdrr.io/cran/data.table/man/data.table.html), [this](https://medium.com/analytics-vidhya/r-data-table-joins-48f00b46ce29), [this](https://stackoverflow.com/questions/15170741) and [this](https://stackoverflow.com/questions/13493124).  To merge any number of datasets at once and obtain a printed report of how the merge went, use the `Hmisc` `Merge` function.]{.aside}

```{r two}
b <- data.table(id=1:4, age=c(21, 28, 32, 23), key='id')
L <- data.table(id  = c(2, 2, 2, 3, 3, 3, 3, 4, 4, 5, 5, 5),
                day = c(1, 2, 3, 1, 2, 3, 4, 1, 2, 1, 2, 3),
                y    =  1 : 12, key='id')
b
L
# Merge b and L to look up baseline age and associate it with all follow-ups
b[L, on=.(id)]   # Keeps all ids in L (left inner join)
L[b, on=.(id)]   # Keeps all ids in b (right inner join)
L[b, on=.(id), nomatch=NULL]  # Keeps only ids in both b and L (right outer join)
uid <- unique(c(b[, id], L[, id]))
L[b[.(uid), on=.(id)]]         # Keeps ids in either b or c (full outer join)
merge(b, L, by='id', all=TRUE) # also full outer join; calls merge.data.table
```

For very large data tables, giving the data tables _keys_ will speed execution, e.g.:

```
setkey(d, id)
setkey(d, state, city)
```

Join/merge can be used for data lookups:

```{r lookup}
s <- data.table(st=.q(AL, AK, AZ, CA, OK), y=5:1)
stateAbbrevs <- data.table(state=state.abb, State=state.name)
s[stateAbbrevs, , on=.(st=state), nomatch=NULL]
```

# Reshaping Data

To reshape data from long to wide format, take the `L` data table above as an example.

```{r l2w}
w <- dcast(L, id ~ day, value.var='y')
w
# Now reverse the procedure
m <- melt(w, id.vars='id', variable.name='day', value.name='y')
m
setkey(m, id, day)   # sorts
m[! is.na(y)]
```

# Other Manipulations of Longitudinal Data

For the longitudinal data table `L` carry forward to 4 days the subject's last observation on `y` if it was assessed earlier than day 4.

```{r}
w <- copy(L)   # fresh start with no propagation of changes back to L
# Only needed if will be using := to compute variables in-place and
# you don't want the new variables also added to L
# This is related to data.table doing things by reference instead of 
# making copies.  w <- L does not create new memory for w.
# Compute each day's within-subject record number and last record number
# Feed this directly into a data.table operation to save last records 
# when the last is on a day < 4
u <- w[, .q(seq, maxseq) := .(1 : .N, .N), by=id][seq == maxseq & day < 4,]
# Extra observations to fill out to day 4
u <- u[, .(day = (day + 1) : 4, y = y), by=id]
u
w <- rbind(L, u, fill=TRUE)
setkey(w, id, day)  # sort and index
w
```

Find the first time at which y >= 3 and at which y >= 7.  [`day[y >= 3]` is read as "the value of `day` when `y >= 3`".  It is a standard subscripting operation in R for two parallel vectors `day` and `y`.  Taking the minimum value of `day` satisfying the condition gives us the first qualifying day.]{.aside}

```{r firstt}
L[, .(first3 = min(day[y >= 3]),
      first7 = min(day[y >= 7])), by=id]
```

Same but instead of resulting in an infinite value if no observations for a subject meet the condition, make the result `NA`.

```{r firstna}
mn <- function(x) if(length(x)) min(x) else as.double(NA)
# as.double needed because day is stored as double precision
# (type contents(L) to see this) and data.table requires
# consistent storage types
L[, .(first3 = mn(day[y >= 3]),
      first7 = mn(day[y >= 7])), by=id]
```

Add a new variable `z` and compute the first day at which `z` is above 0.5 for two days in a row for the subject.  Note that the logic below looks for consecutive days _for which records exist for a subject_.  To also require the days to be one day apart add the clause `day == shift(day) + 1` after `shift(z) > 0.5`.

```{r}
set.seed(1)
w <- copy(L)
w[, z := round(runif(.N), 3)]
u <- copy(w)
u
mn <- function(x)
  if(! length(x) || all(is.na(x))) as.double(NA) else min(x, na.rm=TRUE)
u[, consecutive := z > 0.5 & shift(z) > 0.5, by=id][, 
    firstday    := mn(day[consecutive]),     by=id]
u
```

In general, using methods that involve counters makes logic more clear, easier to incrementally debug, and easier to extend the condition to any number of consecutive times.  Create a function that computes the number of consecutive `TRUE` values or ones such that whenever the sequence is interrupted by `FALSE` or 0 the counting starts over.
As before we compute the first `day` at which two consecutive `z` values exceed 0.5. [`nconsec` is modified from code found [here](https://stackoverflow.com/questions/19998836).]{.aside}


```{r nconsec}
nconsec <- function(x) x * sequence(rle(x)$lengths)
# rle in base R: run length encoding
# Example:
x <- c(0,0,1,1,0,1,1,0,1,1,1,1)
nconsec(x)
# To require that the time gap between measurements must be <= 2 time
# units use the following example
t <- c(1:9, 11, 14, 15)
rbind(t=t, x=x)
nconsec(x & t <= shift(t) + 2)

u <- copy(w)
# nconsec(z > 0.5) = number of consecutive days (counting current
# day) for which the subject had z > 0.5
u[, firstday := mn(day[nconsec(z > 0.5) == 2]), by=id] 
#               |  |  |                    |
#         minimum  |  |                    |
#                day  |                    |
#             such that                    |
#  it's the 2nd consecutive day with z > 0.5
u
```

## Overlap Joins

The `foverlaps` function in `data.table` provides an amazingly fast way to do complex overlap joins.  Our first example is modified from an example in the help file for `foverlaps`.  An annotation column is added to describe what happened.

```{r overlap}
d1 <- data.table(w     =.q(a, a, b, b, b),
                 start = c( 5, 10, 1, 25, 50),
                 end   = c(11, 20, 4, 52, 60))
d2 <- data.table(w     =.q(a, a, b),
                 start = c(1, 15,  1),
                 end   = c(4, 18, 55),
                 name  = .q(dog, cat, giraffe),
                 key   = .q(w, start, end))
f <- foverlaps(d1, d2, type="any")
ann <- c('no a overlap with d1 5-11 & d2 interval',
         'a 10-20 overlaps with a 16-18',
         'b 1-4 overlaps with b 1-55',
         'b 25-62 overlaps with b 1-55',
         'b 50-60 overlaps with b 1-55')
f[, annotation := ann]
f
# Don't include record for non-match
foverlaps(d1, d2, type='any', nomatch=NULL)
# Require the d1 interval to be within the d2 interval
foverlaps(d1, d2, type="within")
# Require the intervals to have the same starting point
foverlaps(d1, d2, type="start")
```

Now consider an example where there is an "events" dataset `e` with 0 or more rows per
subject containing start (`s`) and end (`e`) times and a measurement `x`
representing a daily dose of something given to the subject from `s` to `e`.
The base dataset `b` has one record per subject with times `c` and `d`.  Compute the total dose of drug received between `c` and `d` for the
subject.  This is done by finding all records in `e` for the subject
such that the interval `[c,d]` has any overlap with the interval `[s,e]`.
For each match compute the number of days in the interval `[s,e]` that are
also in `[c,d]`.  This is given by `min(e,d) + 1 - max(c,s)`.  Multiply this
duration by `x` to get the total dose given in `[c,d]`.  For multiple records
with intervals touching `[c,d]` add these products.

```{r totaldose}
base   <- data.table(id    = .q(a,b,c), low=10, hi=20)
events <- data.table(id    = .q(a,b,b,b,k),
                     start = c( 8,  7, 12, 19, 99),
                     end   = c( 9,  8, 14, 88, 99),
                     dose  = c(13, 17, 19, 23, 29))
setkey(base,   id, low,   hi)
setkey(events, id, start, end)
w <- foverlaps(base, events,
               by.x = .q(id, low,   hi),
               by.y = .q(id, start, end ),
               type = 'any', mult='all', nomatch=NA)

w[, elapsed := pmin(end, hi) + 1 - pmax(start, low)]
w[, .(total.dose = sum(dose * elapsed, na.rm=TRUE)), by=id]
```

Similar things are can be done with _non-equi merges_.  For those you can require exact subject matches but allow inexact matches on other variables.  The following example is modified from [here](https://www.scitilab.com/post_data/non_equi_joins/2020_11_17_non_equi_merge). A `medication` dataset holds the start and end dates for a patient being on a treatment, and a second dataset `visit` has one record per subject ID per doctor visit.  For each visit look up the drug in effect if there was one.

```{r noneq}
medication <-
  data.table(ID         = c( 1, 1, 2, 3, 3),
             medication = .q(a, b, a, a, b),
             start      = as.Date(c("2003-03-25","2006-04-27","2008-12-05",
                                    "2004-01-03","2005-09-18")),
             stop       = as.Date(c("2006-04-02","2012-02-03","2011-05-03",
                                    "2005-06-30","2010-07-12")),
             key        = 'ID')
medication
set.seed(123)
visit <- data.table(
  ID   = rep(1:3, 4),
  date = sample(seq(as.Date('2003-01-01'), as.Date('2013-01-01'), 1), 12),
  sbp  = round(rnorm(12, 120, 15)),
  key  = c('ID', 'date'))
visit
# Variables named in inequalities need to have variables in
# medication listed first
m <- medication[visit, on = .(ID, start <= date, stop > date)]
m
# start and stop dates are replaced with actual date of visit
# drop one of them and rename the other
m[, stop := NULL]
setnames(m, 'start', 'date')
m
```

# Graphics

I make heavy use of [ggplot2](https://ggplot2.tidyverse.org), [plotly](https://plotly.com/r) and [R base graphics](https://bookdown.org/rdpeng/exdata/the-base-plotting-system-1.html).  `plotly` is used for interactive graphics, and the [R plotly](https://cran.r-project.org/web/packages/plotly/index.html) package provides an amazing function `ggplotly` to convert a static `ggplot2` graphics object to an interactive `plotly` one.  If the user goes to the trouble of adding labels for graphics entities (usually points, lines, curves, rectangles, and circles) those labels can become hover text in `plotly` without disturbing anything in static graphics.[With `ggplotly` extra text appears in front of labels, but the result of `ggplotly` can be run through `Hmisc::ggplotlyr` to remove this as shown in the example.]{.aside}  As shown [here](https://hbiostat.org/R/simMarkov/sim.html) you can sense whether an `html` or `pdf` report is being produced, and for `html` all `ggplot2` objects can be automatically transformed to `plotly`.

Many types of graphs can be created with base graphics, e.g. `hist(age, nclass=50)` or `Ecdf(age)` but using `ggplot2` for even simple graphics makes it easy to add handle multiple groups on one graph or to create multiple panels for strata using faceting.  `ggplot2` has excellent default font sizes and axis labeling that works for most sizes of plots.

Here is a prototypical `ggplot2` example illustrating many of the features I most often use.  Ignore the `ggplot2` `label` attribute if not using `plotly`.  Options are given to the `Hmisc` `label` function so that it will retrieve the variable `label` and `units` (if present) and format them for axis labels or tables.  The formatting takes into account whether html output is being created and `plotly` is being used.

```{r ggplot2ex}
# Create a vector of formatted labels for all variables in data
# For variables without labels or units use the variable name
# as the label.  If html and plotly are not in effect use R's
# regular plotmath notation to typeset labels/units
nam   <- names(d)
nv    <- length(nam)
vlabs <- structure(character(nv), names=nam)
for(n in nam)
  vlabs[n] <- label(d[[n]], plot=TRUE, html=ishtml, default=n)

# Define substitutes for xlab and ylab that look up our
# constructed labels.
# Could instead directly use xlab(vlabs['age'])
labx <- function(v) xlab(vlabs[[as.character(substitute(v))]])
laby <- function(v) ylab(vlabs[[as.character(substitute(v))]])
g <-
  ggplot(d, aes(x=age, y=bhr, color=gender, label=paste0('dose:', dose))) +
         geom_point() + geom_smooth() +
         guides(color=guide_legend(title='')) +
         theme(legend.position='bottom') +  # not respected by ggplotly
         labs(caption='Scatterplot of age by basal heart rate stratified by sex') +
         labx(age) + laby(bhr)
# or just xlab('Age in years') + ylab('Basal heart rate')
ggplotlyr(g, remove='.*): ')  # removes paste0("dose:", dose): 
# dose is in hover text for each point
```

For large datasets the `Hmisc` package has a function `ggfreqScatter` that makes it easy to see overlapping points by color coding the frequency of points in each small bin.  That way scatterplots scale to very large datasets.  Here is an example:[`html=TRUE` was needed because otherwise axis labels are formatted using R's `plotmath` and `plotly` doesn't like that.]{.aside}

```{r ggfreq}
set.seed(1)
x <- round(rnorm(2000), 1)
y <- 2 * (x > 1.5) + round(rnorm(2000), 1)
z <- sample(c('a', 'b'), 2000, replace=TRUE)
label(x) <- 'X Variable'   # could use xlab() &
label(y) <- 'Y Variable'   # ylab() in ggfreqScatter()
g <- ggfreqScatter(x, y, by=z, html=ishtml)
# If variables were inside a data table use
# g <- d[, ggfreqScatter(x, y, by=z, html=ishtml)]
g
```

Now convert the graphic to `plotly` if html is in effect otherwise stay with `ggplot2` output.

```{r ggfreqp}
ggp(g)
```

When you hover the mouse over a point, its frequency pops up.

Many functions in the `Hmisc` and `rms` packages produce `plotly` graphics directly.  One of the most unique functions is [dotchartpl](https://www.rdocumentation.org/packages/Hmisc/versions/4.6-0/topics/dotchartpl).

# Analysis

## Big Picture

For analysis the sky is the limit, but [statistical principles](https://www.fharrell.com/post/principles) should guide every step.  Some of the general principles are

* If there is to be a pivotal analysis there should be a statistical analysis plan (SAP) for this analysis that does not allow for many "statistician degrees of freedom."  The plan should be completed before doing any analysis that might inform analysis choices in a way that would bias the results (e.g., bias the estimate of treatment effect or bias standard errors of effects in a model).
* All analyses should be completely [reproducible](https://hbiostat.org/bbr/md/repro.html).  Explicitly state random number seeds if any random processes (bootstrap, simulation, Bayesian posterior sampling) are involved.
* Exploratory analysis can take place after any needed SAP is completed.
* [Stay close to the raw data](https://stackoverflow.blog/2022/03/03/stop-aggregating-away-the-signal-in-your-data)
* Continuous or ordinal variables should never be dichotomized even for purely descriptive exploratory analysis.  For example, computing proportions of patients with disease stratified by quintiles of weight will be both inefficient and misleading.
* If the study is a randomized trial, presenting descriptive statistics stratified by treatment ("Table 1") [is not helpful](https://discourse.datamethods.org/t/should-we-ignore-covariate-imbalance-and-stop-presenting-a-stratified-table-one-for-randomized-trials), and it is more informative to describe the overall distribution of subjects.  Even more helpful is to show how all baseline variables relate to the **outcome** variable.
* Above the tendency to interchange the roles of independent and dependent variables by presenting a "Table 2" in such a way that stratifies by the outcome.  Stratifying (conditioning) on the outcome is placing it in the role of a baseline variable.  Instead, show relationships of baseline variables to outcomes as mentioned in the previous point.
* Nonparametric smoothers and estimating in overlapping moving windows are excellent tools for relating individual continuous variables to an outcome.
* Models are often the best descriptive tools because they can account for multiple variables simultaneously.  For example, instead of computing proportions of missing values of a variable Y stratified by age groups and sex, use a binary logistic regression model to relate smooth nonlinear age and sex to the probability Y is missing.

## Replacement for Table 1

Analyses should shed light on the unknown and not dwell on the known.  In a randomized trial, the distributions of baseline variables are expected to be the same across treatments, and will be the same once $N$ is large.  When _apparent_ imbalances are found, they lead to inappropriate decisions and ignore the fact that [apparently counterbalancing factors are not hard to find](https://discourse.datamethods.org/t/should-we-ignore-covariate-imbalance-and-stop-presenting-a-stratified-table-one-for-randomized-trials).  What is unknown and new is how the subject characteristics (and treatment) relate to the outcomes under study.  While displaying this trend with a nonparametric smoother, one can simultaneously display the marginal distribution of the characteristic using an extended box plot, spike histogram, or rug plot.  Here is an example using the `Hmisc` `summaryRc` function.  Extended box plots show the mean, median, and quantiles that cover 0.25, 0.5, 0.75, and 0.9 of the distribution.[For more examples see [this](http://hbiostat.org/R/Hmisc/summaryFuns.pdf)]{.aside}

```{r summaryrc}
#| layout-ncol: 2
getHdata(support)
summaryRc(hospdead ~ age + crea + meanbp + wblc,
          ylim=c(.1, .6), data=support, bpplot='top')
```

Let's take a closer look at extended box plots.  Here is an example in which the distribution of continuous variables is shown, stratified by disease group.

```{r bpplotM}
bpplotM(age + crea + meanbp + wblc ~ dzgroup,
        data=support, cex.strip=0.4, cex.means=0.3, cex.n=0.45)
```

This is better done with interactive plots so that one can for example hover over a corner of a box plot and see which quantile that corner represents.

```{r bpplotMp}
s <- summaryM(age + crea + meanbp + wblc ~ dzgroup,
              data=support)
options(grType='plotly')
plot(s)
```

## Descriptively Relating One Variable to Another

To understand the relationship between a continuous variable X and an outcome or another variable Y we may estimate the mean, median, and other quantities as a smooth function of X[For binary Y the mean is the proportion of ones, which estimates the probability that Y=1]{.aside}.  There are many ways to do this, including

* making a scatter plot if Y is continuous or almost continuous
* stratifying by fixed or variable intervals of X, e.g., summarizing Y by quintiles of X.  This is [arbitrary, inefficient, and misleading](https://discourse.datamethods.org/t/categorizing-continuous-variables) and should never be done.
* using a nonparametric smoother such as `loess`
* parametrically estimating the mean Y as a function of X using an ordinary linear least squares (OLS) model with a regression spline in X so as to not assume linearity
* likewise but with a logistic regression model if Y is binary
* semiparametrically estimating quantiles of Y as a function of X using quantile regression and a regression spline for X
* semiparametrically estimating the mean, quantiles, and exceedance probabilities of Y as a function of X using an ordinal regression model and a spline in X
* nonparametrically using overlapping moving windows of X that advance by a small amount each time.  For each window compute the estimate of the property of Y using ordinary sample estimators (means, quantiles, Kaplan-Meier estimates, etc.).  This approach has the fewest assumptions and is very general in the sense that all types of Y are accommodated.  The moving estimates need to be smoothed; the R `supsmu` function is well suited for this.

The estimated trend curves depend on the window width and amount of smoothing, but this problem is tiny in comparison with the huge effect of changing how a continuous predictor is binned when the usual non-overlapping strata are created.  The idea is assume smooth relationships and get close to the data.

In the following several of the above methods are illustrated to study how serum creatinine of critically ill patients relates to age.  Start with a scatterplot that has no problems with ties in the data.

```{r ggfcr}
with(support, ggfreqScatter(age, crea))
```

Now consider moving estimates, least squares (OLS), ordinal regression (ORM), and quantile regression (QR) estimates, and nonparametric `loess` estimates.  `movStats` provides two methods for creating moving overlapping windows.  The default used here creates varying-width intervals in the data space but fixed-width in terms of sample size.  It includes by default 75 observations to the left of the target point and 75 to the right, and moves up $\max(\frac{n}{200}, 1)$ observations for each evaluation of the statistics.  These may be overridden by specifying `eps` and `xinc`.  The mean and all three quartiles are estimated for each window.  All this is obtained by `movStats`, which is [on Github](https://github.com/harrelfe/rscripts/blob/master/movStats.r) and makes heavy use of the `data.table` and `rms` packages.  For ordinal regression estimates of the mean and quantiles the log-log link is used in the example below.  Moving estimates are shown with and without [`supsmu`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/supsmu)-smoothing them.

```{r moving}
#| fig.height: 5
#| fig.width: 7
getRs('movStats.r', put='source')
u <- movStats(crea ~ age,
              loess=TRUE, ols=TRUE, qreg=TRUE,
              orm=TRUE, family='loglog', msmooth='both',
              melt=TRUE, data=support)
ggplot(u, aes(x=age, y=crea, col=Type)) + geom_line() +
  facet_wrap(~ Statistic) +
  xlab('Age') + ylab('Serum Creatinine') +
  labs(caption=attr(u, 'info'))
 ```

Recommended practice for relating a continuous variable to another continuous variable, especially for replacing parts of Table 1 or Table 2, is to use smoothed moving statistics or (1) a spline OLS model to estimate the mean and (2) a spline quantile regression model for estimating quantiles.  Here is an example best practice that shows a preferred subset of the estimates from the last plot. `melt=TRUE` is omitted so we can draw a ribbon to depict the outer quartiles.

```{r creapref}
u <- movStats(crea ~ age, data=support)
ggplot(u, aes(x=age, y=`Moving Median`)) + geom_line() +
  geom_ribbon(aes(ymin=`Moving Q1`, ymax=`Moving Q3`), alpha=0.2) +
  geom_line(aes(x=age, y=`Moving Mean`, col=I('blue'))) +
  xlab('Age') + ylab('Serum Creatinine') +
  labs(caption='Black line: median\nBlue line: mean\nBand: Q1 & Q3')
```

`r cwbc <- '$\\sqrt[3]{\\mathrm{WBC}}$'`
Let's describe how white blood count relates to the probability of hospital death, using a binary logistic regression model and moving proportions.
The cube root transformation in regression fits is used because of the extreme skewness of WBC.  Use 6 knots at default locations on `r cwbc`.  The `r cwbc` transformation affects moving statistics only in that mean x-values for plotting are cubes of mean `r cwbc` instead of means on the original WBC scale.

```{r wbc}
u <- movStats(hospdead ~ wblc, k=6,
              trans  = function(x) x ^ (1/3),
              itrans = function(x) x ^ 3,
              loess=TRUE, lrm=TRUE, msmooth='both',
              melt=TRUE, data=support)
ggplot(u, aes(x=wblc, y=hospdead, col=Type)) + geom_line() +
  guides(color=guide_legend(title='')) +
  theme(legend.position='bottom',
        plot.caption=element_text(family='mono', size=6)) +
  labs(caption=attr(u, 'info'))
```

The flexibility of the moving statistic method is demonstrated by estimating how age relates to probabilities of death within 1y and within 2y using Kaplan-Meier estimates in overlapping moving windows.  Assumptions other than smoothness (e.g., proportional hazards) are avoided in this approach.  Here is an example that also uses an flexible parametric method, [hazard regression](https://www.tandfonline.com/doi/abs/10.1080/01621459.1995.10476491), implemented in the R [`polspline`](https://cran.r-project.org/web/packages/polspline/index.html) package, that adaptively finds knots (points of slope change) in the covariate and in time, and products of piecewise linear terms so as to allow for non-proportional hazards.  We use far less penalization than is the default for the `hare` function for demonstration purposes.  For this dataset the default settings of `penalty` and `maxdim` result in straight lines.

```{r km}
u <- movStats(Surv(d.time / 365.25, death) ~ age, times=1:2,
              hare=TRUE, penalty=0.5, maxdim=30,
              melt=TRUE, data=support)
ggplot(u, aes(x=age, y=incidence, col=Statistic)) + geom_line() +
  facet_wrap(~ Type) +
  ylab(label(u$incidence)) +
  guides(color=guide_legend(title='')) +
  theme(legend.position='bottom')
```

## One Continuous and One Categorical Predictor

It is possible to descriptively estimate trends against more than one independent variables when the effective sample size is sufficient.  Trends can be estimated nonparametrically through stratification (when the third variable is categorical) or with flexible regression models allowing the two predictors to interact.  In the graphical displays it is useful to keep sample size limitations in certain regions of the space defined by the two predictors in mind, by superimposing spike histograms on trend curves.

Repeat the last example but stratified by disease class.  The window is widened a bit because of the reduced sample size upon stratification.  Default smoothing is used for hazard regression.

```{r km2}
# The Coma stratum has only n=60 so is not compatible with eps=75
# Use varyeps options
u <- movStats(Surv(d.time / 365.25, death) ~ age + dzclass, times=1:2,
              varyeps=TRUE, msmooth='both', hare=TRUE,
              melt=TRUE, data=support)
ggplot(u, aes(x=age, y=incidence, col=dzclass)) + geom_line() +
  facet_grid(Type ~ Statistic) +
  ylab(label(u$incidence)) +
  guides(color=guide_legend(title='')) +
  theme(legend.position='bottom',
        plot.caption=element_text(family='mono', size=6)) +
  labs(caption=attr(u, 'info'))
```

Consider another example with a continuous dependent variable.  Use the NHANES dataset that was created for analyzing glycohemoglobin (HbA$_{\mathrm{1c}}$) for diabetes screening.  Stratify by race/ethnicity

```{r nhanes}
getHdata(nhgh)
u <- movStats(gh ~ age + re,
              melt=TRUE, data=nhgh)
ggplot(u, aes(x=age, y=gh, col=re)) + geom_line() +
  facet_wrap( ~ Statistic) +
  ylab(label(nhgh$gh)) +
  guides(color=guide_legend(title='', nrow=2)) +
  theme(legend.position='bottom',
        plot.caption=element_text(family='mono', size=6)) +
  labs(caption=attr(u, 'info'))

```

Mimic these results using flexible regression with interaction.  Start by estimating the mean.  Add spike histograms to estimated trend curves.  Spike heights are proportional to the sample size in age/race-ethnicity groups after binning age into 100 bins.  Direct `plotly` plotting is used.  The user can click on elements of the legend (including the histograms) to turn their display off and on.

```{r ghols,results='asis'}
require(rms)
options(prType='html')  # needed to use special formatting (can use prType='latex')
dd <- datadist(nhgh); options(datadist='dd')
f <- ols(gh ~ rcs(age, 5) * re, data=nhgh)
anova(f)
hso <- list(frac=function(f) 0.1 * f / max(f),
            side=1, nint=100)
# Plot with plotly directly
plotp(Predict(f, age, re), rdata=nhgh, histSpike.opts=hso)
```

Now use quantile regression to estimate quartiles of glycohemoglobin as a function of age and race/ethnicity.

```{r nhghq}
f1 <- Rq(gh ~ rcs(age, 5) * re, tau=0.25, data=nhgh)
f2 <- Rq(gh ~ rcs(age, 5) * re, tau=0.5,  data=nhgh)
f3 <- Rq(gh ~ rcs(age, 5) * re, tau=0.75, data=nhgh)
p  <- rbind(Q1     = Predict(f1, age, re, conf.int=FALSE),
            Median = Predict(f2, age, re, conf.int=FALSE),
            Q3     = Predict(f3, age, re, conf.int=FALSE))
ggplot(p, histSpike.opts=hso)
```

## Formatting

I take advantage of special formatting for model fit objects from the `rms` package by using `html` or `latex` methods and putting `results='asis'` in the chunk header to preserve the formatting.  Here is an example.

```{r lrm,results='asis'}
require(rms)
options(prType='html')  # needed to use special formatting (can use prType='latex')
set.seed(1)
n <- 1000
w <- data.table(x=runif(n), x2=rnorm(n), y=sample(0:1, n, replace=TRUE))
dd <- datadist(w); options(datadist='dd')  # rms needs for summaries and plotting
f <- lrm(y ~ x + rcs(x2, 4), data=w)
f
anova(f)
latex(f)   # show fitted model in math notation
```

## Caching

The workhorse behind `Rmarkdown` and `Quarto` (besides [`Pandoc`](https://pandoc.org)) is [knitr](https://www.r-project.org/nosvn/pandoc/knitr.html), which processes the code chunks and properly mingles code and tabular and graphical output.  `knitr` has a built-in caching mechanism to make it so that code is not needlessly executed when the code inputs have not changed. This easy-to-use process does have two disadvantages: the dependencies are not transparent, and the stored cache files may be quite large.  I like to take control of caching.  To that end, the [runifChanged](https://github.com/harrelfe/rscripts/blob/master/hashCheck.r) function was written.  Here is an example of its use.  First a function with no arguments must be composed.  This is the (usually slow) function that will be conditionally run if any of a group of listed objects has changed since the last time it was run.  This function when needed to be run produces an object that is stored in binary form in a user-specified file (the default file name is the name of the current R code chunk with `.rds` appended).

```{r runifch}
# Read the source code for the hashCheck and runifChanged functions from
# https://github.com/harrelfe/rscripts/blob/master/hashCheck.r
getRs('hashCheck.r', put='source')
g <- function() {
  # Fit a logistic regression model and bootstrap it 500 times, saving
  # the matrix of bootstrapped coefficients
  f <- lrm(y ~ x1 + x2, x=TRUE, y=TRUE, data=dat)
  bootcov(f, B=500)
}
set.seed(3)
n   <- 2000
dat <- data.table(x1=runif(n), x2=runif(n),
                  y=sample(0:1, n, replace=TRUE))
# runifChanged will write runifch.rds if needed (chunk name.rds)
# Will run if dat or source code for lrm or bootcov change
b <- runifChanged(g, dat, lrm, bootcov)
dim(b$boot.Coef)
head(b$boot.Coef)
```

## Parallel Computing

The `runParallel` function makes it easy to use available multiprocessor cores to speed up parallel computations especially for simulations.  By default it runs the number of available cores, less one.  `runParallel` makes the `parallel` package easier to use and does recombinations over per-core batches.  The user writes a function that does the work on one core, and the same function is run on all cores.  This function has set arguments and must return a named list.  A base random number `seed` is given, and the seed is set to this plus `i` for core number `i`.  The total number of repetitions is given, and this most balanced possible number of repetitions is run on each core to sum to the total desired number of iterations.  `runifChanged` is again used, to avoid running the simulations if no inputs have changed.

```{r rflowpar}
# Load runParallel function from github
getRs('runParallel.r', put='source')

# Function to do simulations on one core
run1 <- function(reps, showprogress, core) {
  cof <- matrix(NA, nrow=reps, ncol=3,
                dimnames=list(NULL, .q(a, b1, b2)))
  for(i in 1 : reps) {
    y <- sample(0:1, n, replace=TRUE)
    f <- lrm(y ~ X)
    cof[i, ] <- coef(f)
  }
  list(coef=cof)
}
# Debug one core run, with only 3 iterations
n    <- 300
seed <- 3
set.seed(seed)
X    <- cbind(x1=runif(n), x2=runif(n))  # condition on covariates
run1(3)

nsim <- 5000
g <- function() runParallel(run1, reps=nsim, seed=seed)
Coefs <- runifChanged(g, X, run1, nsim, seed)
dim(Coefs)
apply(Coefs, 2, mean)
```

# Simulation

Some of the best ways to validate an analysis are

* If using any model/feature selection methods use the bootstrap to check whether the selection process is volatile, e.g., your sample size isn't large enough too support making hard-and-fast selections of predictors/features
* Use Monte Carlo simulation to check if the correct model or correct predictors are usually selected
* Simulate a large dataset under a known model and known parameter values and make sure the estimation process you use can recover the true parameter values
* Simulate the statistical performance of a method under a variety of conditions

Unlike static papers in the literature, simulation can study the performance of a method under conditions that mimic your situation.

When simulating performance of various quantities under various conditions, creating a large number of variables makes the code long and tedious.  It is better to to use data frames/tables or arrays to hold everything together.  Data frames and arrays also lead to efficient graphics code for summarization.

## Data Table Approach

The `expand.grid` function is useful for generating all combinations of simulation conditions.  Suppose we wanted to simulate statistical properties of the maximum absolute value of the sample correlation coefficient from a matrix of all pairwise correlations from truly uncorrelated variables.  We do this while varying the sample size `n`, the number of variables `p`, and the `type` of correlation (Pearson's or Spearman's, denoted by `r` and `rho`). With `expand.grid` we don't need a lot of nested `for` loops.  Run 500 simulations for each condition.

```{r simd}
nsim <- 500
R <- expand.grid(n=c(10, 20, 50, 100),
                 p=c(2, 5, 10, 20),
                 sim=1 : nsim)
setDT(R)
set.seed(1)
for(i in 1 : nrow(R)) {  # took 4s
  w <- R[i, ]
  n <- w$n
  p <- w$p
  X <- matrix(rnorm(n * p), ncol=p)
  cors    <- cor(X)
  maxr    <- max(abs(cors[row(cors) < col(cors)])) # use upper triangle
  cors    <- cor(X, method='spearman')
  maxrho  <- max(abs(cors[row(cors) < col(cors)]))
  set(R, i, 'maxr',   maxr)    # set is in data.table & is very fast
  set(R, i, 'maxrho', maxrho)  # set will create the variable if needed
  # If not using data.table use this slower approach:
  # R[i, 'maxr'] <- maxr   etc.
}
```

The simulations could have been cached or parallelized as discussed above.

Compute the mean (over simulations) maximum correlation (over variables) and plot the results.

```{r simdpp}
w <- R[, .(maxr = mean(maxr), maxrho=mean(maxrho)), by=.(n, p)]
# Make data table taller and thinner to put r, rho as different observations
u <- melt(w, id.vars=c('n', 'p'), variable.name='type', value.name='r')
u[, type := substring(type, 4)]   # remove "max"
ps <- c(2, 5, 10, 20)
u[, p := factor(p, ps, paste0('p:', ps))]
g <- ggplot(u, aes(x=n, y=r, col=type)) + geom_jitter(height=0, width=2) +
      ylim(0, 1) +
      facet_wrap(~ p) +
      guides(color=guide_legend(title='')) +
      ylab('Mean Maximum Correlation Coefficient')
plotly::ggplotly(g)
```

## Array Approach

For large problems, storing results in R `array`s is more efficient and doesn't require duplication of values of `n` and `p` over simulations.  Once the `array` is created it can be converted into a data table for graphing.

```{r sima}
nsim <- 500
ns   <- c(10, 20, 50, 100)
ps   <- c(2, 5, 10, 20)
R <- array(NA, dim=c(nsim, length(ns), length(ps), 2),
               dimnames=list(NULL,
                             n    = as.character(ns),
                             p    = as.character(ps),
                             type = c('r', 'rho')))
dim(R)
dimnames(R)
set.seed(1)
for(i in 1 : nsim) {   # took 1s
  for(n in ns) {
    for(p in ps) {
      X <- matrix(rnorm(n * p), ncol=p)
      for(type in c('r', 'rho')) {
        cors <- cor(X, method=switch(type, r = 'pearson', rho = 'spearman'))
        # or           method=fcase(type == 'r', 'pearson', type == 'rho', 'spearman')
        # or           method=fcase(type == 'r', 'pearson', default='spearman')
        # or           method=c(r='pearson', rho='spearman')[type]
        # or           method=.q(r=pearson, rho=spearman)[type]
        # or           method=if(type == 'r') 'pearson' else 'spearman'
        # or           method=ifelse(type == 'r', 'pearson', 'spearman')
        maxr <- max(abs(cors[row(cors) < col(cors)]))
        R[i, as.character(n), as.character(p), type] <- maxr
      }
    }
  }
}
# Compute mean (over simulations) maximum correlation for each condition
m <- apply(R, 2:4, mean)   # preserve dimensions 2,3,4 summarize over 1
# Convert the 3-dimensional array to a tall and thin data table
# Generalizations of row() and col() used for 2-dimensional matrices
# comes in handy: slice.index
dn <- dimnames(m)
u <- data.table(r    = as.vector(m),
                n    = as.numeric(dn[[1]])[as.vector(slice.index(m, 1))],
                p    = as.numeric(dn[[2]])[as.vector(slice.index(m, 2))],
                type = dn[[3]][as.vector(slice.index(m, 3))])
# If doing this a lot you may want to write a dimension expander function
slice <- function(a, i) {
  dn <- all.is.numeric(dimnames(a)[[i]], 'vector')   # all.is.numeric in Hmisc
  dn[as.vector(slice.index(a, i))]
}
u <- data.table(r    = as.vector(m),
                n    = slice(m, 1),
                p    = slice(m, 2),
                type = slice(m, 3))
  
# Plot u using same ggplot code as above
```

# Other Resources

* [Software development resources for data scientists](https://www.rstudio.com/blog/software-development-resources-for-data-scientists) by ISABELLA VELSQUEZ

# Computing Environment

The following output is created by the command `markupSpecs$html$session()`, where `markupSpecs` is defined in the `Hmisc` package.

`r markupSpecs$html$session()`
